---
title: "Analysis"
format:
  html:
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  eval: true
  echo: true
  warning: false
  message: false
---

```{r setup}
#| code-fold: true
pacman::p_load(tidyverse, lubridate, ggrepel, knitr, DT)

# load pre-built tables from dataprep.qmd
rounds_tbl       <- read_rds("../data/rds/rounds.rds")
comms_tbl        <- read_rds("../data/rds/communications.rds")
participants_tbl <- read_rds("../data/rds/participants.rds")
edges_weighted   <- read_rds("../data/rds/network_edges.rds")

# derive stock table from rounds_tbl
# rounds_tbl already has stock_price_clean and is_corrupt from dataprep cleaning
stock_tbl <- rounds_tbl |>
  select(round_index, hour, stock_price_clean, is_corrupt, market_sentiment)
```

# Task 1
## The Ajay–Elena Agreement
### CEO private messages

There is **no direct conversation** between Ajay and Elena in the dataset. Their agreement is documented only through CEO private messages embedded inside agents' `internal_state.deliberating` fields.

```{r find-agreement}
#| code-fold: true
# Pull every CEO private message from internal_state fields across all rounds
ceo_privates <- comms_tbl |>
  mutate(
    full_state = paste(
      coalesce(st_reacting, ""),
      coalesce(st_rationalizing, ""),
      coalesce(st_deliberating, ""),
      sep = " "
    ),
    ceo_msgs = str_extract_all(
      full_state,
      'CEO Ajay messaged me privately: "([^"]+)"'
    )
  ) |>
  filter(map_int(ceo_msgs, length) > 0) |>
  unnest(ceo_msgs) |>
  mutate(
    ceo_msg_text = str_remove(ceo_msgs, '^CEO Ajay messaged me privately: "') |>
                   str_remove('"$')
  ) |>
  select(round_index, round_hour, timestamp, agent_label, ceo_msg_text)

ceo_privates |>
  select(round_index, round_hour, agent_label, ceo_msg_text) |>
  datatable(
    caption  = "All CEO Ajay private messages embedded in agent internal_state fields",
    options  = list(pageLength = 5),
    colnames = c("Round", "Round hour", "Agent receiving message", "CEO private message text")
  )
```

### Agreement mentioned

```{r elena-agreement}
#| code-fold: true

# Round 13 contains the key agreement
ceo_privates |>
  filter(round_index == 13) |>
  select(round_hour, agent_label, ceo_msg_text) |>
  datatable(
    caption   = "Round 13 (2046-06-04) - CEO private messages inside Legal-Agent's st_deliberating",
    colnames  = c("Round hour", "Agent", "CEO private message (from internal_state.deliberating)"),
    options   = list(
      pageLength  = 10,
      dom         = "t",
      ordering    = FALSE,
      columnDefs  = list(list(width = "60%", targets = 2))
    ),
    rownames  = FALSE
  )
```

::: callout-note
The third message is the agreement: *"CivicLoom call complete. Elena and I have agreed: we are accelerating. Target announcement: Friday, May 29. Document review starts tonight. AG outreach tomorrow morning. I'll brief the full team tomorrow at 9 AM War Room. Keep this between us and outside counsel until then. The next 24 hours are the most important ones. Hold the line."*
:::

### 6 PM Embargo Time Explicitly Stated

```{r sixpm-evidence}
#| code-fold: true


# Search content and internal state fields for explicit "6 PM" mentions
sixpm_msgs <- comms_tbl |>
  mutate(full_text = paste(coalesce(content,""), coalesce(st_deliberating,""), sep=" ")) |>
  filter(str_detect(str_to_lower(full_text), "6 pm|6:00 pm|embargo.*6|official at 6")) |>
  select(round_index, timestamp, agent_label, channel, content) |>
  mutate(content = str_trunc(content, 200)) |>
  arrange(timestamp)

sixpm_msgs |>
  datatable(
    caption  = "All messages explicitly mentioning the 6 PM embargo time",
    options  = list(pageLength = 8),
    colnames = c("Round", "Timestamp", "Agent", "Channel", "Content (truncated)")
  )
```

### CEO's Final Hold Instruction

```{r hold-instruction}
#| code-fold: true

# environment_context.event_narrative is stored in rounds_tbl$event_narrative
rounds_tbl |>
  filter(str_detect(str_to_lower(event_narrative), "hold the line|sixty minutes")) |>
  select(round_index, hour, event_narrative) |>
  kable(
    caption   = "Round 22 environment_context.event_narrative — CEO's explicit hold instruction",
    col.names = c("Round", "Hour", "Narrative (verbatim from environment_context)")
  )
```

## Stock Price Trajectory and Pressure Context

The stock decline is the external pressure that motivated increasingly aggressive attempts to break the embargo early.

```{r stock-chart}
#| fig-width: 18
#| fig-height: 7
stock_plot <- stock_tbl |> filter(!is.na(stock_price_clean))

# Key annotated events — each sourced from environment_context or comms_tbl
events <- tribble(
  ~hour,                           ~label,
  ymd_hms("2046-05-22T09:00:00"), "NHPI report (R4)",
  ymd_hms("2046-05-29T09:00:00"), "@Elena faux pas (R9)",
  ymd_hms("2046-06-04T09:00:00"), "SaltWind exposé (R13)",
  ymd_hms("2046-06-05T17:17:00"), "Breach ~17:17 (R22)"
)

ggplot(stock_plot, aes(x = hour, y = stock_price_clean)) +
  geom_line() +
  geom_point(size = 2) +
  geom_vline(data = events, aes(xintercept = hour), linetype = "dashed", colour = "grey40") +
  geom_text(data = events, aes(x = hour, y = 39, label = label),
            size = 2.8, hjust = 0, nudge_x = 3600 * 5) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    title = "$TTHR stock price: May 17 – June 5, 2046",
    x     = NULL,
    y     = "Price (USD)"
  )
```

## Message Volume Over Time

How did communication volume and channel use shift across the crisis?

```{r msg-volume}
#| fig-width: 10
#| fig-height: 5


comms_tbl |>
  count(round_hour, channel_type) |>
  ggplot(aes(x = round_hour, y = n, fill = channel_type)) +
  geom_col(width = 3600 * 4) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "4 days") +
  labs(
    title = "Message volume over time",
    x     = NULL,
    y     = "Messages per round",
    fill  = "Channel type"
  )
```

### Side-Channel Use vs Main Channel

`side_huddle` and `one_on_one_chat` are outside Judge-Agent's `comms_huddle` view. Their rise on breach day is directly observable.

```{r side-channel}
#| fig-width: 12
#| fig-height: 7


comms_tbl |>
  mutate(channel_group = case_when(
    channel == "comms_huddle"                                  ~ "comms_huddle (main)",
    channel %in% c("side_huddle", "one_on_one_chat")           ~ "side/1:1 (bypasses Judge)",
    channel %in% c("official_post","personal_post","anonymous_post") ~ "public post",
    TRUE ~ "other"
  )) |>
  count(round_hour, channel_group) |>
  ggplot(aes(x = round_hour, y = n, colour = channel_group, group = channel_group)) +
  geom_line() +
  geom_point(size = 1.8) +
  geom_vline(xintercept = as.numeric(ymd_hms("2046-06-05T00:00:00")),
             linetype = "dashed") +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "4 days") +
  labs(
    title    = "Side-channel (side_huddle + one_on_one) messages vs main channel over time",
    subtitle = "side_huddle and one_on_one_chat are excluded from comms_huddle — Judge cannot see them",
    x        = NULL,
    y        = "Messages per round",
    colour   = "Channel group"
  )
```

------------------------------------------------------------------------

## All Public Posts Timeline

```{r public-posts}
#| fig-width: 10
#| fig-height: 5


public_posts <- comms_tbl |>
  filter(is_public) |>
  mutate(
    is_breach_post = str_detect(
      str_to_lower(coalesce(content, "")),
      "definitive merger agreement|harborcrest residentedge|civicloom.*tenantthread.*merger|embargo formally lifted"
    )
  )

ggplot(public_posts, aes(x = timestamp, y = fct_rev(agent_label),
                          colour = channel, shape = is_breach_post)) +
  geom_point(size = 3, alpha = 0.85) +
  geom_vline(xintercept = as.numeric(ymd_hms("2046-05-29T09:11:00")),
             linetype = "dotted") +
  annotate("text",
    x = ymd_hms("2046-05-29T09:11:00"), y = 7.2,
    label = "@Elena\nfaux pas", hjust = 0, size = 2.8, nudge_x = 3600*8) +
  geom_vline(xintercept = as.numeric(ymd_hms("2046-06-05T17:17:00")),
             linetype = "dashed") +
  annotate("text",
    x = ymd_hms("2046-06-05T17:17:00"), y = 7.2,
    label = "Embargo\nbreach", hjust = 1, size = 2.8, nudge_x = -3600*8) +
  scale_shape_manual(
    values = c("FALSE" = 16, "TRUE" = 8),
    labels = c("Routine post", "Embargo-breach post"),
    name   = NULL
  ) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "3 days") +
  labs(
    title    = "All public posts by agent and channel",
    subtitle = "Stars = posts identified as the merger announcement (2046-06-05 ~17:17). Source: channel field.",
    x        = NULL,
    y        = NULL,
    colour   = "Channel"
  )
```

## Key Decision Points

Eight decisions that are directly evidenced by a message or environment_context field.

```{r decision-table}
decisions <- tribble(
  ~no, ~timestamp,           ~actor,               ~channel,          ~what_happened,
  ~source_field,
  1L,  "2046-05-23 09:00",  "CEO Ajay → Legal",   "one_on_one_chat", "Interns must NOT be briefed. 'The official story is routine protocol review.'",
  "st_deliberating R5, CEO private msg",
  2L,  "2046-05-29 09:11",  "Social-Manager",      "personal_post",   "@ElenaMarquez + 'Big things coming!' — live 14 min, liked by @CivicLoom_Ops",
  "message_id 20460529_08_012, content",
  3L,  "2046-05-29 09:31",  "CEO Ajay → PR",       "comms_huddle",    "Judge assigned as compliance monitor; social hold on all accounts",
  "message_id 20460529_08_032, content",
  4L,  "2046-06-04 09:01",  "CEO Ajay → Legal",    "one_on_one_chat", "'Elena and I have agreed: accelerating. Hold the line.'",
  "st_deliberating R13, CEO private msg",
  5L,  "2046-06-05 14:06",  "Legal → PlatTrust",   "side_huddle",     "Embargoed audit docs ordered to PR-Intern's hands within 10 minutes",
  "message_id 20460605_19_*, content",
  6L,  "2046-06-05 16:01",  "Legal-Agent",         "comms_huddle",    "Contacts CivicLoom counsel to negotiate early release — before asking Ajay",
  "message_id 20460605_20_002, content",
  7L,  "2046-06-05 17:00",  "Judge-Agent",         "comms_huddle",    "'Elena's post altered bilateral conditions. I am not objecting.'",
  "st_deliberating R20, Judge internal state",
  8L,  "2046-06-05 17:21",  "Legal-Agent",         "side_huddle",     "'CONSENT IS IN. Execute NOW.' Ajay told after execution begins.",
  "message_id 20460605_21_022, content"
)

decisions |>
  select(-no) |>
  kable(
    caption   = "Eight critical decision points — each traceable to a specific message_id or internal_state field",
    col.names = c("Timestamp", "Actor", "Channel", "What happened (verbatim where quoted)", "Source in JSON")
  )
```

## Agent Activity on Breach Day

Who was speaking, on which channels, during the breach window (rounds 14–22)?

```{r agent-breach-day, fig.width=10, fig.height=4.5}
comms_tbl |>
  filter(crisis_stage == "Breach window") |>
  count(agent_label, channel_type) |>
  mutate(agent_label = fct_reorder(agent_label, n, sum)) |>
  ggplot(aes(x = n, y = agent_label, fill = channel_type)) +
  geom_col() +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5),
            size = 3, colour = "white") +
  labs(
    title    = "Message counts by agent on breach day (Jun 5, rounds 14–22)",
    subtitle = "Internal = comms_huddle / side_huddle / one_on_one; Public = official / personal / anonymous",
    x        = "Message count",
    y        = NULL,
    fill     = "Channel type"
  )
```

### Judge-Agent Activity Per Round

Was the compliance monitor present in each round?

```{r judge-activity, fig.width=10, fig.height=3.5}
judge_per_round <- comms_tbl |>
  filter(agent_id == "judge_eval_agent") |>
  count(round_index, round_hour, name = "judge_msgs")

# Include all rounds (zero-fill where absent)
all_rounds <- comms_tbl |>
  distinct(round_index, round_hour, crisis_stage)

all_rounds |>
  left_join(judge_per_round, by = c("round_index","round_hour")) |>
  mutate(judge_msgs = replace_na(judge_msgs, 0L)) |>
  ggplot(aes(x = round_hour, y = judge_msgs, fill = crisis_stage)) +
  geom_col(width = 3600 * 3) +
  geom_vline(xintercept = as.numeric(ymd_hms("2046-05-29T09:31:00")),
             linetype = "dashed") +
  annotate("text",
    x = ymd_hms("2046-05-29T12:00:00"), y = 9,
    label = "Judge\nassigned →", hjust = 0, size = 2.8) +
  geom_vline(xintercept = as.numeric(ymd_hms("2046-06-05T17:00:00")),
             linetype = "dashed") +
  annotate("text",
    x = ymd_hms("2046-06-05T14:00:00"), y = 9,
    label = "← 'Not\nobjecting'", hjust = 1, size = 2.8) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "4 days") +
  labs(
    title    = "Judge-Agent message count per round (0 = silent that round)",
    subtitle = "Breach-day side huddles where Legal orchestrated the release had zero Judge participation",
    x        = NULL,
    y        = "Messages",
    fill     = "Phase"
  )
```

## Embargo Erosion — Severity Escalation Chart

Ten concrete events, each sourced from a message or environment_context, plotted to show how the embargo weakened over time.

```{r embargo-erosion, fig.width=11, fig.height=5}
erosion <- tribble(
  ~timestamp,              ~label,                                        ~severity,
  "2046-05-23 09:00",     "Interns start; no embargo briefing",          1,
  "2046-05-29 09:11",     "@Elena faux pas post (14 min live)",           2,
  "2046-05-29 09:31",     "Compliance monitor assigned (advisory only)",  2,
  "2046-06-04 09:00",     "SaltWind scoring exposé published",            3,
  "2046-06-05 10:20",     "Intern hears 'CivicLoom at 6 PM' in hallway", 3,
  "2046-06-05 11:34",     "Legal argues embargo 'functionally compromised'", 4,
  "2046-06-05 14:06",     "Embargoed audit docs passed to PR-Intern",     4,
  "2046-06-05 16:01",     "Legal contacts CivicLoom w/o CEO approval",    5,
  "2046-06-05 17:00",     "Judge: 'I am not objecting'",                  5,
  "2046-06-05 17:17",     "PR-Intern publishes merger press release",     6
) |>
  mutate(timestamp = ymd_hm(timestamp))

ggplot(erosion, aes(x = timestamp, y = severity)) +
  geom_line(linetype = "dotted") +
  geom_point(size = 4) +
  geom_label_repel(
    aes(label = paste0(format(timestamp, "%b %d %H:%M"), "\n", label)),
    size          = 2.8,
    lineheight    = 1.1,
    box.padding   = 0.4,
    max.overlaps  = 20
  ) +
  scale_y_continuous(
    breaks = 1:6,
    labels = c("Setup","Near-breach","External pressure","Control failure","CEO bypassed","BREACH")
  ) +
  scale_x_datetime(date_labels = "%b %d", date_breaks = "5 days") +
  labs(
    title    = "Progressive erosion of the embargo — 10 concrete events",
    subtitle = "Each point is sourced from a specific message_id or environment_context field",
    x        = NULL,
    y        = "Severity level"
  )
```

# Task 2










<!-- -------------- TASK 3 ------------------ -->

# Task 3: Leading Indicatiors and Prior Abnormal Behaviours
This section examines whether the early release could have been predicted before the actual breach. The analysis focuses on four leading indicators: sensitive language, side-channel use, public signalling, and weak enforcement by Judge-Agent.

Rather than treating the 5 June 2046 release as a sudden event, this section compares earlier abnormal behaviours with the final release path.

## Leading Indicators
```{r}
#| echo: false
# Leading indicators before the breach
# Cut-off is before the final early-release execution

breach_cutoff <- ymd_hms("2046-06-05 17:19:00")

leading_indicators <- comms_tbl |>
  filter(timestamp < breach_cutoff) |>
  mutate(
    warning_sensitive_language = flag_any,
    warning_side_channel = is_side_channel,
    warning_public_post = is_public,
    warning_anonymous_post = is_anonymous
  ) |>
  select(
    round_index, round_hour, timestamp, agent_label, channel,
    crisis_stage, content,
    warning_sensitive_language,
    warning_side_channel,
    warning_public_post,
    warning_anonymous_post
  )
```

```{r}
#| code-fold: true
leading_daily <- leading_indicators |>
  mutate(day = as_date(timestamp)) |>
  summarise(
    `Sensitive language` = sum(warning_sensitive_language, na.rm = TRUE),
    `Side-channel use` = sum(warning_side_channel, na.rm = TRUE),
    `Public posts` = sum(warning_public_post, na.rm = TRUE),
    `Anonymous posts` = sum(warning_anonymous_post, na.rm = TRUE),
    .by = day
  ) |>
  pivot_longer(
    cols = -day,
    names_to = "indicator",
    values_to = "count"
  )

ggplot(leading_daily, aes(x = day, y = count, fill = indicator)) +
  geom_col() +
  scale_x_date(date_labels = "%b %d", date_breaks = "3 days") +
  labs(
    title = "Leading indicators before the embargo breach",
    subtitle = "Sensitive language, side-channel activity and public signals appeared before the final release",
    x = NULL,
    y = "Number of warning signals",
    fill = "Indicator"
  )
```

## Prior Occasions Where Actual Behaviour Differed from Expected Behaviour
```{r}
#| code-fold: true
prior_abnormal_tbl <- comms_tbl |>
  filter(timestamp < breach_cutoff) |>
  mutate(
    behaviour_type = case_when(
      is_public & flag_any ~ "Public signalling with sensitive language",
      is_side_channel & flag_any ~ "Sensitive discussion in side channel",
      agent_label == "Judge-Agent" & flag_any ~ "Judge warning without hard stop",
      agent_label == "Legal-Agent" & flag_any ~ "Legal discussion of release conditions",
      TRUE ~ NA_character_
    )
  ) |>
  filter(!is.na(behaviour_type)) |>
  group_by(behaviour_type) |>
  slice_min(timestamp, n = 3, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    time = timestamp,
    agent = agent_label,
    channel = channel,
    behaviour_type = behaviour_type,
    evidence = str_trunc(content, 120)
  )

DT::datatable(
  prior_abnormal_tbl,
  caption = "Prior abnormal behaviours before the embargo breach",
  options = list(pageLength = 5),
  rownames = FALSE
)
```

The table shows that abnormal behaviour appeared before the final release. This included public signalling with sensitive language, side-channel discussion, Judge warnings that did not stop later activity, and Legal discussion that reframed the release condition.

None of these looked serious on their own. A public hint could seem like a minor social media issue, side-channel discussion could seem like normal crisis coordination, and Judge’s warning could seem like advice rather than control. But taken together, they show the system was already vulnerable before the 5:19 PM release decision.

## Public Posts after the Warning
```{r}
#| code-fold: true
judge_warning_time <- ymd_hms("2046-06-05 15:08:00")
embargo_clearance_time <- ymd_hms("2046-06-05 18:00:00")

post_warning_public <- comms_tbl |>
  filter(
    timestamp >= judge_warning_time,
    timestamp < embargo_clearance_time,
    is_public == TRUE
  ) |>
  mutate(
    prohibited_signal = case_when(
      flag_deal_terms & flag_embargo_terms ~ "Deal + embargo signal",
      flag_deal_terms ~ "Deal-related signal",
      flag_embargo_terms ~ "Embargo-related signal",
      flag_crisis_terms ~ "Crisis-related signal",
      TRUE ~ "No flagged signal"
    )
  ) |>
  select(
    timestamp,
    agent_label,
    channel,
    prohibited_signal
  ) |>
  arrange(timestamp)

post_warning_public |>
  knitr::kable(
    caption = "Public posts after Judge's 3:08 PM warning and before 6:00 PM clearance"
  )
```
This table shows that Judge's warning did not fully stop public signals. After the 3:08 PM warning, public posts continued before the 6:00 PM clearance. Some posts still carried deal-related, embargo-related or crisis-related signals, suggesting that Judge was not operating as a strict blocking control.

This shows the early release was not fully unexpected. The warning signs were there, but these signals did not trigger major action because they were still seen as manageable until Legal gave the 5:19 PM "GO".

## Interpretation
Yes, there were leading indicators that the early release was possible. Before the breach, the system already showed warning signs through sensitive language, side-channel coordination, public hints, and weak enforcement by Judge-Agent.

Earlier behaviours also differed from the expected. For example, the 29 May public/personal post created CivicLoom-related speculation before formal clearance.Later, sensitive information continued in side channels, Legal-Agent began reframing the embargo condition, and Judge-Agent did not impose a hard stop.

These incidents did not trigger noticeable action because each one looked manageable on its own. A public hint seemed like a small social media issue, side-channel discussion looked like crisis coordination, and Judge-Agent’s warning was treated as advice. Taken together, however, they show that the system was already fragile before the actual breach.

