---
title: "Data Preparation"
format:
  html:
    toc-depth: 3
    number-sections: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

This document walks through how we parse, clean, and reshape the raw VAST Challenge 2026 Mini-Challenge 1 dataset into the three relational tables, which are **Rounds**, **Communications**, and **Participants**.  

# Setup

## Loading Packages

We use the following packages:

- [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) - parse the JSON source file
- [tidyverse](https://www.tidyverse.org/) - data wrangling and string operations
- [lubridate](https://lubridate.tidyverse.org/) - timestamp parsing
- [tidytext](https://cran.r-project.org/web/packages/tidytext/index.html) - tokenisation for keyword flagging and text analysis
- [knitr](https://cran.r-project.org/web/packages/knitr/index.html) - formatted table output

```{r}
pacman::p_load(jsonlite, tidyverse, lubridate, tidytext, knitr)
```

## Loading Data

We load the single JSON file provided for VAST Challenge 2026 Mini-Challenge 1.

```{r}
mc1_raw <- fromJSON("data/MC1_final_00.json", simplifyVector = FALSE)
rounds_raw <- mc1_raw$rounds
cat("Total rounds:", length(rounds_raw), "\n")
```

The dataset contains **23 rounds**, spanning 17 May 2046 to 5 June 2046 (the day of the embargo breach). Each round holds an `environment_context` object, a flat list of `communications`, and a `participants` list recording each agent's status for that round.

# Understanding the Data Structure

## Top-level schema

Each round has four keys:

| Key | Description |
|---|---|
| `hour` | Timestamp for the round start |
| `environment_context` | External conditions: market data, events, headlines |
| `communications` | All messages sent in that round |
| `participants` | Agent metadata and declared actions |


## Agents

| Agent ID               | Role           | Label          | Seniority          | Description                                                                                                                    |
| ---------------------- | -------------- | -------------- | ------------------ | ------------------------------------------------------------------------------------------------------------------------------ |
| `legal_agent`          | legal          | Legal-Agent    | Senior             | General legal counsel.                                                                                                         |
| `quality_agent`        | platform_trust | Platform-Trust | Senior             | VP of Platform Trust & Safety.                                                                                                 |
| `social_manager_agent` | social_manager | Social-Manager | Senior             |Manages Social messaging         |
| `pr_agent`             | pr             | PR-Agent       | Senior             | Head of Communications and Public Relationships |
| `intern_agent`         | intern         | Intern         | Junior             | General intern.                                                                                 |
| `pr_intern_agent`      | pr_intern      | PR-Intern      | Junior             | PR team intern with access to the official TenantThread Flex account.                                                          |
| `judge_eval_agent`     | judge          | Judge          | Compliance officer | Evaluates risks, mediates conflicts, provides compliance guidance.                                                             |

## Communication channels and message types

```{r}
#| code-fold: true

# Flatten all communications to inspect channels and message types
all_comms_raw <- map_dfr(rounds_raw, function(rnd) {
  map_dfr(rnd$communications, function(msg) {
    tibble(
      channel      = msg$channel      %||% NA_character_,
      message_type = msg$message_type %||% NA_character_
    )
  })
})

all_comms_raw %>%
  count(channel, message_type, name = "count") %>%
  arrange(desc(count)) %>%
  kable()
```

Channels fall into two broad groups:

- **Internal**: `comms_huddle`, `side_huddle`, `one_on_one_chat`
- **Public**: `official_post`, `personal_post`, `anonymous_post`


# Assigning Crisis Stages

Before extracting tables, we assign each round to one of four analytical windows. Rounds 1-13 cover the pre-crisis and crisis build-up, rounds 14-22 cover the breach day, round 23 is the post-embargo period.

```{r}
#| code-fold: true

crisis_stage_map <- tibble(
  round_index  = 1:23,
  crisis_stage = c(
    # Pre-crisis: 17 May - 24 May (rounds 1-6)
    rep("Pre-crisis", 6),
    # Crisis pre-breach: 25 May - 4 June (rounds 7-13)
    rep("Crisis (pre-breach)", 7),
    # Breach window: 5 June 09:00–17:00 (rounds 14-22)
    rep("Breach window", 9),
    # Post-embargo: 5 June 18:00 (round 23)
    "Post-embargo"
  )
) %>%
  mutate(crisis_stage = factor(
    crisis_stage,
    levels = c("Pre-crisis", "Crisis (pre-breach)", "Breach window", "Post-embargo")
  ))

kable(crisis_stage_map %>% count(crisis_stage))
```

# Preparing the Rounds Table

## Extracting fields

We extract one row per round from `environment_context`, including the (corrupted) raw stock price for cleaning in section @sec-data-cleaning. 

```{r}
#| code-fold: true

rounds_tbl <- map_dfr(seq_along(rounds_raw), function(i) {
  rnd <- rounds_raw[[i]]
  ctx <- rnd$environment_context
  ms  <- ctx$market_snapshot

  tibble(
    round_index        = i,
    hour               = ymd_hms(rnd$hour),
    event_headline     = ctx$event_headline   %||% NA_character_,
    event_narrative    = ctx$event_narrative  %||% NA_character_,
    social_state       = ctx$social_state     %||% NA_character_,
    stock_price_raw    = ms$stock_price       %||% NA_character_,
    pct_change_raw     = ms$percent_change    %||% NA_character_,
    market_sentiment   = ms$sentiment         %||% NA_character_,
    critical_deadlines = paste(unlist(ctx$critical_deadlines %||% list()),
                               collapse = "; "),
    news               = paste(unlist(ctx$news %||% list()),
                               collapse = "; "),
    agents_unavailable = paste(unlist(ctx$agents_unavailable %||% list()),
                               collapse = ", ")
  )
}) %>%
  left_join(crisis_stage_map, by = "round_index")

glimpse(rounds_tbl)
```

## Cleaning corrupted stock prices {#sec-data-cleaning}

::: {.callout-warning title="Known Data Quality Issues"}
- `event_headline` for rounds 15 and 16 both read `"10:00 AM."`, corrected using the `hour` timestamp instead.
- `stock_price` for several crisis-day rounds contains corrupted values (`$180`, `NULL`), replaced with prices quoted in agent messages.
:::

```{r}
#| code-fold: true

# Prices recovered from agent message content for each corrupt round
stock_price_fixes <- tribble(
  ~round_index, ~stock_price_clean,
  14L,  28.70,    # 09:00 — NULL, opening price from agent content
  15L,  31.50,    # 10:00 — NULL, interpolated from agent content
  16L,  NA_real_, # 11:00 — $180 corrupt, no reliable quote found
  19L,  26.90,    # 14:00 — $18.0 corrupt, use 13:00 price as reference
  20L,  32.80,    # 15:00 — $18.0 corrupt, quoted in side-huddle
  21L,  33.05,    # 16:00 — NULL, Sentinel 5:12 PM update
  22L,  33.05,    # 17:00 — NULL, same post-market reference
  23L,  33.05     # 18:00 — NULL, after-hours Sentinel final report
)

rounds_tbl <- rounds_tbl %>%
  select(-any_of(c("stock_price_raw_num", "is_corrupt",
                   "stock_price_clean", "pct_change_clean"))) %>%
  mutate(
    stock_price_raw_num = as.numeric(str_remove(stock_price_raw, "\\$")),
    is_corrupt = round_index %in% stock_price_fixes$round_index |
                 stock_price_raw_num %in% c(18, 18.0, 180) |
                 is.na(stock_price_raw)
  ) %>%
  left_join(stock_price_fixes, by = "round_index") %>%
  mutate(
    stock_price_clean = if_else(
      !is_corrupt,
      stock_price_raw_num,
      stock_price_clean
    ),
    pct_change_clean = as.numeric(str_remove(pct_change_raw, "%"))
  )
```



# Preparing the Communications Table

## Extracting fields

We flatten all messages from all rounds into a single data frame. Internal state sub-fields (`reacting`, `rationalizing`, `deliberating`) are extracted as separate columns.

```{r}
#| code-fold: true

comms_tbl <- map_dfr(seq_along(rounds_raw), function(i) {
  rnd  <- rounds_raw[[i]]
  hour <- ymd_hms(rnd$hour)

  map_dfr(rnd$communications, function(msg) {
    ist <- msg$internal_state %||% list()

    tibble(
      round_index    = i,
      round_hour     = hour,
      message_id     = msg$message_id     %||% NA_character_,
      agent_id       = msg$agent_id       %||% NA_character_,
      agent_role     = msg$agent_role     %||% NA_character_,
      agent_label    = msg$agent_label    %||% NA_character_,
      channel        = msg$channel        %||% NA_character_,
      message_type   = msg$message_type   %||% NA_character_,
      responding_to  = msg$responding_to  %||% NA_character_,
      content        = msg$content        %||% NA_character_,
      timestamp      = ymd_hms(msg$timestamp %||% NA_character_),
      recipients     = paste(unlist(msg$recipients %||% list()), collapse = ", "),
      # Internal state sub-fields
      st_reacting      = ist$reacting      %||% NA_character_,
      st_rationalizing = ist$rationalizing %||% NA_character_,
      st_deliberating  = ist$deliberating  %||% NA_character_
    )
  })
}) %>%
  left_join(crisis_stage_map, by = "round_index")

cat("Total messages:", nrow(comms_tbl), "\n")
```

## Deriving temporal and structural columns

```{r}
#| code-fold: true

comms_tbl <- comms_tbl %>%
  mutate(
    date              = as_date(timestamp),
    hour_of_day       = hour(timestamp),
    day_of_week       = wday(timestamp, label = TRUE),
    is_crisis_day     = date == as_date("2046-06-05"),
    # Minutes elapsed since first message in dataset (for continuous timeline)
    minutes_since_start = as.numeric(difftime(timestamp,
                                               min(timestamp, na.rm = TRUE),
                                               units = "mins")),
    # Channel groupings
    channel_type      = case_when(
      channel %in% c("comms_huddle", "side_huddle", "one_on_one_chat") ~ "Internal",
      channel %in% c("official_post", "personal_post", "anonymous_post") ~ "Public",
      TRUE ~ "Other"
    ),
    is_side_channel   = channel %in% c("side_huddle", "one_on_one_chat"),
    is_public         = channel %in% c("official_post", "personal_post", "anonymous_post"),
    is_anonymous      = channel == "anonymous_post",
    is_broadcast      = recipients == "ALL",
    recipient_count   = if_else(
      is_broadcast,
      NA_integer_,
      str_count(recipients, ",") + 1L
    ),
    has_internal_state = !is.na(st_reacting) | !is.na(st_rationalizing) |
                         !is.na(st_deliberating)
  )
```


## Keyword flagging

We create three Boolean flag columns on `content` and (where present) `st_deliberating`, covering deal-related, embargo-related, and crisis-related terms. A combined text field `full_text` is used for this scan.

```{r}
#| code-fold: true

deal_terms    <- c("merger", "acquisition", "deal", "civicloom", "harborcrest",
                   "counterparty", "nda", "term sheet", "due diligence", "signing",
                   "strategic", "transaction", "partner")

embargo_terms <- c("embargo", "hold", "silence", "6 pm", "6:00", "release",
                   "clearance", "judge", "compliance", "unauthorized", "approved",
                   "restricted", "monitor")

crisis_terms  <- c("breach", "leak", "premature", "crisis", "firestorm",
                   "escalate", "saltwind", "oceancrunch", "ag inquiry",
                   "investigation", "viral", "senate", "attorney general")

flag_text <- function(text_col, terms) {
  pattern <- paste(terms, collapse = "|")
  str_detect(str_to_lower(text_col), pattern)
}

comms_tbl <- comms_tbl %>%
  mutate(
    full_text          = paste(coalesce(content, ""),
                                coalesce(st_deliberating, ""), sep = " "),
    flag_deal_terms    = flag_text(full_text, deal_terms),
    flag_embargo_terms = flag_text(full_text, embargo_terms),
    flag_crisis_terms  = flag_text(full_text, crisis_terms),
    flag_any           = flag_deal_terms | flag_embargo_terms | flag_crisis_terms
  )

DT::datatable(comms_tbl, caption = "Communications Table", options = list(pageLength = 5),class = "nowrap")
```


## Back-filling comm_count into rounds table

```{r}
#| code-fold: true

comm_counts <- comms_tbl %>%
  count(round_index, name = "comm_count")

rounds_tbl <- rounds_tbl %>%
  select(-any_of("comm_count")) %>%
  left_join(comm_counts, by = "round_index")

DT::datatable(rounds_tbl, caption = "Rounds Table", options = list(pageLength = 5),class = "nowrap")
```


# Preparing the Participants Table

## Extracting fields

We extract one row per agent per round from the `participants` list, then parse `declared_action` into a clean action type and a separate content field.

```{r}
#| code-fold: true

participants_tbl <- map_dfr(seq_along(rounds_raw), function(i) {
  rnd <- rounds_raw[[i]]
  map_dfr(rnd$participants %||% list(), function(p) {
    meta <- p$agent_round_metadata %||% list()
    tibble(
      round_index         = i,
      round_hour          = ymd_hms(rnd$hour),
      agent_id            = p$agent_id            %||% NA_character_,
      agent_role          = p$agent_role           %||% NA_character_,
      agent_label         = p$agent_label          %||% NA_character_,
      declared_action_raw = p$declared_action      %||% NA_character_,
      sentiment_at_turn   = meta$sentiment_at_turn %||% NA_character_,
      action_classification = meta$action_classification %||% NA_character_
    )
  })
}) %>%
  left_join(crisis_stage_map, by = "round_index")
```


## Parsing declared action type

`declared_action` can be a short token (e.g. `MONITORING`) or a long string containing post content. We extract the action type prefix and store the remainder as `action_content`.

```{r}
#| code-fold: true

participants_tbl <- participants_tbl %>%
  mutate(
    action_type = case_when(
      is.na(declared_action_raw)                            ~ NA_character_,
      str_starts(declared_action_raw, "MONITORING")         ~ "MONITORING",
      str_starts(declared_action_raw, "POSTED_ON_FLEX")     ~ "POSTED_ON_FLEX",
      str_starts(declared_action_raw, "POSTED_ANONYMOUS")   ~ "POSTED_ANONYMOUS",
      str_starts(declared_action_raw, "POSTED_PERSONAL")    ~ "POSTED_PERSONAL",
      str_starts(declared_action_raw, "COMPLIANCE_WARNING") ~ "COMPLIANCE_WARNING",
      TRUE                                                  ~ "OTHER"
    ),
    action_content = str_remove(declared_action_raw,
                                "^(MONITORING|POSTED_ON_FLEX|POSTED_ANONYMOUS|POSTED_PERSONAL|COMPLIANCE_WARNING)[:\\s—]*")
  )

DT::datatable(participants_tbl, options = list(pageLength = 5),class = "nowrap")
```


# Building the Network Edge List

For network analysis we need a directed sender-to-recipient edge list. Where `recipients = "ALL"`, we expand to one edge per other agent active in that round.

## Expanding broadcast messages

Direct messages between agents. 

```{r}
#| code-fold: true

# Active agents per round (from participants table)
active_agents_per_round <- participants_tbl %>%
  select(round_index, agent_id) %>%
  distinct()

# Separate broadcast vs direct messages
direct_msgs <- comms_tbl %>%
  filter(!is_broadcast, !is.na(recipients), recipients != "") %>%
  separate_rows(recipients, sep = ",\\s*") %>%
  select(message_id, round_index, crisis_stage, timestamp, channel,
         sender = agent_id, recipient = recipients)

DT::datatable(direct_msgs, options = list(pageLength = 5),class = "nowrap")
```

Messages broadcasted among agents. 

```{r}
#| code-fold: true

broadcast_msgs <- comms_tbl %>%
  filter(is_broadcast) %>%
  select(message_id, round_index, crisis_stage, timestamp, channel,
         sender = agent_id) %>%
  left_join(active_agents_per_round, by = "round_index") %>%
  rename(recipient = agent_id) %>%
  filter(recipient != sender)

DT::datatable(broadcast_msgs, options = list(pageLength = 5),class = "nowrap")

edges_tbl <- bind_rows(direct_msgs, broadcast_msgs) %>%
  # Remove messages with no identified recipient
  filter(!is.na(recipient), recipient != "") %>%
  # Exclude internal-state-only rows (no actual sender)
  filter(!is.na(sender))

cat("Total directed edges:", nrow(edges_tbl), "\n")
```


## Adding edge weights by crisis stage

```{r}
edges_weighted <- edges_tbl %>%
  group_by(sender, recipient, channel, crisis_stage) %>%
  summarise(
    weight         = n(),
    first_contact  = min(timestamp, na.rm = TRUE),
    last_contact   = max(timestamp, na.rm = TRUE),
    .groups        = "drop"
  )

DT::datatable(edges_weighted, options = list(pageLength = 5),class = "nowrap")
```















