---
title: "Data Preparation"
format:
  html:
    toc-depth: 3
    number-sections: true
execute:
  eval: true
  echo: true
  freeze: true
  warning: false
  message: false
---

This document walks through how we parse, clean, and reshape the raw VAST Challenge 2026 Mini-Challenge 1 dataset into the three relational tables, which are **Rounds**, **Communications**, and **Participants**.  

# 1 Setup

## 1.1 Loading Packages

We use the following packages:

- [jsonlite](https://cran.r-project.org/web/packages/jsonlite/index.html) - parse the JSON source file
- [tidyverse](https://www.tidyverse.org/) - data wrangling and string operations
- [lubridate](https://lubridate.tidyverse.org/) - timestamp parsing
- [tidytext](https://cran.r-project.org/web/packages/tidytext/index.html) - tokenisation for keyword flagging and text analysis
- [knitr](https://cran.r-project.org/web/packages/knitr/index.html) - formatted table output

```{r}
pacman::p_load(jsonlite, tidyverse, lubridate, tidytext, knitr)
```

## 1.2 Loading Data

We load the single JSON file provided for VAST Challenge 2026 Mini-Challenge 1.

```{r}
mc1_raw <- fromJSON("data/MC1_final_00.json", simplifyVector = FALSE)
rounds_raw <- mc1_raw$rounds
cat("Total rounds:", length(rounds_raw), "\n")
```

The dataset contains **23 rounds**, spanning 17 May 2046 to 5 June 2046 (the day of the embargo breach). Each round holds an `environment_context` object, a flat list of `communications`, and a `participants` list recording each agent's status for that round.

# 2 Understanding the Data Structure

## 2.1 Top-level schema

Each round has four keys:

| Key | Description |
|---|---|
| `hour` | Timestamp for the round start |
| `environment_context` | External conditions: market data, events, headlines |
| `communications` | All messages sent in that round |
| `participants` | Agent metadata and declared actions |


## 2.2 Agents

| Agent ID               | Role           | Label          | Seniority          | Description                                                                                                                    |
| ---------------------- | -------------- | -------------- | ------------------ | ------------------------------------------------------------------------------------------------------------------------------ |
| `legal_agent`          | legal          | Legal-Agent    | Senior             | General legal counsel.                                                                                                         |
| `quality_agent`        | platform_trust | Platform-Trust | Senior             | VP of Platform Trust & Safety.                                                                                                 |
| `social_manager_agent` | social_manager | Social-Manager | Senior             |Manages Social messaging         |
| `pr_agent`             | pr             | PR-Agent       | Senior             | Head of Communications and Public Relationships |
| `intern_agent`         | intern         | Intern         | Junior             | General intern.                                                                                 |
| `pr_intern_agent`      | pr_intern      | PR-Intern      | Junior             | PR team intern with access to the official TenantThread Flex account.                                                          |
| `judge_eval_agent`     | judge          | Judge          | Compliance officer | Evaluates risks, mediates conflicts, provides compliance guidance.                                                             |

## 2.3 Communication channels and message types

```{r}
# Flatten all communications to inspect channels and message types
all_comms_raw <- map_dfr(rounds_raw, function(rnd) {
  map_dfr(rnd$communications, function(msg) {
    tibble(
      channel      = msg$channel      %||% NA_character_,
      message_type = msg$message_type %||% NA_character_
    )
  })
})

all_comms_raw %>%
  count(channel, message_type, name = "count") %>%
  arrange(desc(count)) %>%
  kable()
```

Channels fall into two broad groups:

- **Internal**: `comms_huddle`, `side_huddle`, `one_on_one_chat`
- **Public**: `official_post`, `personal_post`, `anonymous_post`











